Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-29 2024pipit/section/20-Motivation.tex
copy

At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
Each basic cycle consists of a sequence of actions to be performed at specific time-marks.
Actions in the schedule may not be relevant to all nodes; the node's \emph{node matrix} contains only the relevant actions.
The node matrix is represented by a constant \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
Watch alerts are generally placed after an expected reference message to signal an error if no reference message is received.

@@

The network has strict timing requirements, which prohibit the driver from looping through the entire triggers array at each time-mark.
Instead, the driver maintains an index that refers to the current trigger.
At each time-mark, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.

@@

Our driver may fall behind when trying to run certain schedules, as the driver only processes one trigger per time-mark.
To ensure that the schedule can be executed, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next-enabled trigger starts.

Recall our concrete triggers array from \autoref{f:tt-systemmatrix-ok}, which contained trigger 1 (SEND B at time-mark 1 on cycle C0), and trigger 2 (SEND C at time-mark 2 on cycle C1).
We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
However, we could \emph{not} bring trigger 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.

>>>>

At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
Each basic cycle consists of a sequence of actions to be performed at specific time-marks.
Actions in the schedule may not be relevant to all nodes; the node's \emph{node matrix} contains only the relevant actions.
The node matrix is represented in memory by a \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
Watch alerts are generally placed after an expected reference message to signal an error if no reference message is received.

@@

The network has strict timing requirements which prohibit the driver from looping through the entire triggers array at each time-mark.
Instead, the driver maintains an index that refers to the current trigger.
At each time-mark, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.

@@

Our driver may fall behind when trying to run certain schedules, as the driver only processes one trigger per time-mark.
To ensure that the schedule can be executed on time, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next enabled trigger starts.

Recall our concrete triggers array from \autoref{f:tt-systemmatrix-ok}, which contained trigger 1 (SEND B at time-mark 1 on cycle C0), and trigger 2 (SEND C at time-mark 2 on cycle C1).
We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
However, we could \emph{not} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.

### 2024-03-29 2024pipit/section/30-Core.tex
copy

Recursive streams are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
% We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
% Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

@@

Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ involve two verification conditions.
Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
Conceptually, then, a contract could have two property statuses: one for the definition and one for the instantiation.
However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers $i$ by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.

@@

At a high level, evaluation unfolds recursive streams to determine a concrete value.
For example, to evaluate the earlier sum example with input $i = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumionetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream to get $(\xfby{0}{\sumrec}) + i$.
Evaluation of primitives is standard.
For variables, we find the variable in the current heap:
$$
\ruleIN{
}{
  \bigstep{\sumionetwo}{i}{2}
}{Var}
$$

For delays, we discard the current heap and continue evaluation with the history prefix:
$$
\ruleIN{
  \bigstep{\sgl{i \mapsto 1}}{\sumrec}{1}
}{
  \bigstep{\sumionetwo}{\xfby{0}{\sumrec}}{1}
}{$\mbox{Fby}_S$}
$$

@@

% To show that an expression $e$'s unknown properties hold, we show that for all streaming histories $\Sigma$, the unknown properties hold ($\semcheck{\Sigma}{\PSUnknown}{e}$).
% When performing this proof, we can assume that the valid properties already hold to avoid re-checking properties ($\semcheck{\Sigma}{\PSValid}{e}$).
To check a property $(\xcheckP{\PStatus}{e})$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.

@@

For example, recall our earlier contract that the sum of strictly positive integers is positive:
$$
@let@~\text{sum}~i = \sumcontract
$$
To check the contract definition on a concrete input $i = [1; 2]$, we first evaluate the body:
$$
\bigsteps{\sumionetwo}{\sumrec}{[1; 3]}
$$
We then check that, assuming all inputs are positive, then all results are positive:
$$
\bigstepalways{\sumionetwo}{i > 0} \implies
\bigstepalways{\sgl{i \mapsto 1, \sumvar \mapsto 1}; \sgl{i \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
$$

It is crucial that the rely is true \emph{at all points} in the stream.
Consider if we had instead used the input stream $i = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.

To check a contract instance, we need to check whether the call-site satisfies the rely.

@@

We generally prove these properties via abstract transition systems (\autoref{s:transition}).

>>>>

Recursive streams are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but the stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
% We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
% Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

@@

Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ allow modular reasoning by replacing the implementation with an abstract specification.
Contracts involve two verification conditions.
Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
Conceptually, then, a contract could have two property statuses: one for the definition and one for the instantiation.
However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers \emph{ints} by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.

@@

At a high level, evaluation unfolds recursive streams to determine a value.
For example, to evaluate the earlier sum example with input $\textit{ints} = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumionetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream one step to get $(\xfby{0}{\sumrec}) + \textit{ints}$.
Evaluation of primitives is standard.
To evaluate variables, we look for the variable in the current heap:
$$
\ruleIN{
}{
  \bigstep{\sumionetwo}{\textit{ints}}{2}
}{Var}
$$

For delays, we discard the current heap and continue evaluation with the history prefix:
$$
\ruleIN{
  \bigstep{\sgl{\textit{ints} \mapsto 1}}{\sumrec}{1}
}{
  \bigstep{\sumionetwo}{\xfby{0}{\sumrec}}{1}
}{$\mbox{Fby}_S$}
$$

@@

% To show that an expression $e$'s unknown properties hold, we show that for all streaming histories $\Sigma$, the unknown properties hold ($\semcheck{\Sigma}{\PSUnknown}{e}$).
% When performing this proof, we can assume that the valid properties already hold to avoid re-checking properties ($\semcheck{\Sigma}{\PSValid}{e}$).
To check a property $(\xcheckP{\PStatus}{e})$ in history $\Sigma$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.

@@

For example, recall our earlier contract that the sum of strictly positive integers is positive:
$$
@let@~\text{sum}~\textit{i} = \sumcontractX{\PSUnknown}{i}
$$
To check the contract definition on a concrete input $\textit{i} = [1; 2]$, we first evaluate the body:
$$
\bigsteps{\sumionetwo}{\sumrecX{i}}{[1; 3]}
$$
We then check that, assuming all inputs are positive, then all results are positive:
$$
\bigstepalways{\sumionetwo}{\textit{i} > 0} \implies 
\bigstepalways{\sgl{\textit{i} \mapsto 1, \sumvar \mapsto 1}; \sgl{\textit{i} \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
$$

It is critical that the rely is true \emph{at all points} in the stream.
Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
The contract itself remains valid, as the input did not satisfy the rely at all steps.
This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.

Checking a contract instance is simple, in comparison: we just check that the call-site satisfies the rely at all steps.

@@

We generally prove the required properties via abstract transition systems (\autoref{s:transition}).
